The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-08-06 paper2_JNLS/paper.tex
NM

Now it is possible to choose the scales $\ampscale, \lengthscale, \timescale$ for ease in working with a particular framework. We choose the following scales,
\begin{align}
\ampscale = \pi^{1/2} \, \etastd \, , \qquad
\lengthscale = \frac{\lamfac \lam}{\pi} \, , \qquad
\timescale = \frac{\lamfac \lam}{\freqp \depth}
\end{align}
where $\lamfac$ is an integer to be chosen later. 
The explanation for these choices is as follows. First, we have chosen the characteristic amplitude, $\ampscale$, above to normalize the energy of the state-variable $u$ to unity, as will be demonstrated in Section \ref{tKdVSec}. Second, regarding $\lengthscale$, recall that $\lam$ is the characteristic wavelength corresponding to the peak forcing frequency $\freqp$ in the experiments. If only integer multiples of $\freqp$ were imposed (e.g.~lower frequencies were not present), then the forcing would produce waves that are periodic over lengthscale $\lam$. Since lower frequencies do exist, strict periodicity is not satisfied, but rather waves may be nearly periodic over the physical domain $\xi \in [-\lam,\lam]$. The approximation of near-periodicity becomes more accurate if integer multiples of $\lam$ are considered, i.e.~$\xi \in [-\lamfac \lam, \lamfac \lam]$. Thus, we have chosen $\lengthscale$ above so that the dimensionless domain of consideration is $\tilde{x} \in [-\pi, \pi]$ and periodic boundary conditions on $u$ can be  imposed on over this domain with accuracy that increases with $\lamfac$. Lastly, regarding $\timescale$, the most basic timescale in the experiments is simply $\freqp^{-1}$, i.e.~the period of waves passing a fixed reference point. Of course, the leading-order behavior in shallow water is simple translation of waves with uniform speed $c$. The KdV equation provides the first correction to this behavior and describes dynamics that evolve over longer timescales. Hence we have rescaled $\freqp^{-1}$ by the factor $N \lam/\depth \gg 1$, which provides a suitably long timescale and is in line with previous normalizations \cite{johnson1997modern}. The scales $\lengthscale$ and $\timescale$ change value across the ADC, which is important to note when comparing the theory against experimental measurements.

@@

The variable-depth KdV \eqref{dimlessKdV}, though not Hamiltonian throughout the entire domain, admits a Hamiltonian structure on each side of the ADC. Indeed, \eqref{dimlessKdV} can be expressed as
\begin{align}
\dot{u}_{\pm} = \sympJ \vard{\Hupdn}{u}
\end{align}
where $\sympJ = \pdi{}{x}$ is the symplectic operator and $\Ham = \Hupdn$ is the Hamiltonian, which will take different forms on either side of the ADC. It is convenient to decompose the Hamiltonian into a so-called cubic and quadratic component, given respectively by
\begin{align}
\Hthree = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
\Htwo = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}
Then the Hamiltonian can be expressed as
\begin{equation}
\Hupdn = C_2 \dratupdn^{1/2} \, \Htwo - C_3 \dratupdn^{-3/2} \, \Hthree
\end{equation}
where $\drat = \dratupdn$ changes value across the ADC. More explicitly, substituting \eqref{dratpw}, the separate upstream and downstream Hamiltonians are
\begin{align}
&\Hup = C_2 \, \Htwo - C_3 \, \Hthree 						&& \text{for } t<0 \\
&\Hdn = C_2 \dratdn^{1/2} \, \Htwo - C_3 \dratdn^{-3/2} \, \Hthree	&& \text{for } t>0
\end{align}
The above formulation differs cosmetically from some previous ones \cite{abramov2003hamiltonian, bajars2013weakly, majda2019statistical}, which used the opposite sign in both $\sympJ$ and $\Ham$. With that sign convention, a {\em negative} inverse temperature is required to accurately describe the experimental observations, as found in \cite{majda2019statistical}. We, however, choose the sign convention above so that a {\em positive} inverse temperature could be used and thus fit the theory into the most conventional form of statistical mechanics.

@@

@@

Consider the Fourier series of the state variable
\begin{equation}
u = \sum_{k=-\Lambda}^{\Lambda} \uhat_k e^{i k x}
\end{equation}
where
\begin{equation}
\uhat_k = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x) e^{-i k x} \dx
\end{equation}
The above sum can be infinite $\Lambda = \infty$ or truncated at a finite wavenumber $\Lambda < \infty$. Further, since $u$ is real valued, we have $\uhat_{-k} = \uhat_{k}^*$.

We measure the state variable $u$ as the displacement of the surface from equilibrium, so that $\mean{u} = 0$. Consequently, the momentum of $u$ vanishes
\begin{equation}
\Mo[u] \equiv \int_{-\pi}^{\pi} u \dx = 0
\end{equation}
Next, the definition of the characteristic amplitude implies that $\mean{u^2} = \pi^{-1}$. Consequently, the energy of $u$ is fixed as
\begin{equation}
\En[u] \equiv \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = 1
\end{equation}
By Parseval's identity, we have
\begin{equation}
\En[u] = 1 = \frac{1}{2} \int u^2 \dx = \pi \sum_{k=-\Lambda}^{\Lambda} \abs{\uhat_k}^2 = 2 \pi \sum_{k=1}^{\Lambda} \abs{\uhat_k}^2
\end{equation}
% The above agrees with the first line on top of p. 4 in Majda and Qi J Stat Phys 2019.

>>>>

Now it is possible to choose the scales $\ampscale, \lengthscale, \timescale$ for ease in working with a particular framework. We choose the following scales,
\begin{align}
\label{scales}
\ampscale = \pi^{1/2} \, \etastd \, , \qquad
\lengthscale = \frac{\lamfac \lam}{\pi} \, , \qquad
\timescale = \frac{\lamfac \lam}{\freqp \depth}
\end{align}
where $\lamfac$ is an integer to be chosen later. 
The explanation for these choices is as follows. First, we have chosen the characteristic amplitude, $\ampscale$, above to normalize the energy of the state-variable $u$ to unity, as will be demonstrated in Section \ref{tKdVSec}. Second, regarding $\lengthscale$, recall that $\lam$ is the characteristic wavelength corresponding to the peak forcing frequency $\freqp$ in the experiments. If only integer multiples of $\freqp$ were imposed (e.g.~lower frequencies were not present), then the forcing would produce waves that are periodic over lengthscale $\lam$. Since lower frequencies do exist, strict periodicity is not satisfied, but rather waves may be nearly periodic over the physical domain $\xi \in [-\lam,\lam]$. The approximation of near-periodicity becomes more accurate if integer multiples of $\lam$ are considered, i.e.~$\xi \in [-\lamfac \lam, \lamfac \lam]$. Thus, we have chosen $\lengthscale$ above so that the dimensionless domain of consideration is $\tilde{x} \in [-\pi, \pi]$ and periodic boundary conditions on $u$ can be  imposed on over this domain with accuracy that increases with $\lamfac$. Lastly, regarding $\timescale$, the most basic timescale in the experiments is simply $\freqp^{-1}$, i.e.~the period of waves passing a fixed reference point. Of course, the leading-order behavior in shallow water is simple translation of waves with uniform speed $c$. The KdV equation provides the first correction to this behavior and describes dynamics that evolve over longer timescales. Hence we have rescaled $\freqp^{-1}$ by the factor $N \lam/\depth \gg 1$, which provides a suitably long timescale and is in line with previous normalizations \cite{johnson1997modern}. The scales $\lengthscale$ and $\timescale$ change value across the ADC, which is important to note when comparing the theory against experimental measurements.

@@

The variable-depth KdV \eqref{dimlessKdV}, though not Hamiltonian throughout the entire domain, admits a Hamiltonian structure on each side of the ADC. Indeed, \eqref{dimlessKdV} can be expressed as
\begin{align}
\dot{u}_{\pm} = \sympJ \vard{\Hupdn}{u}
\end{align}
where $\sympJ = \pdi{}{x}$ is the symplectic operator and $\Ham = \Hupdn$ is the Hamiltonian, which will take different forms on either side of the ADC. It is convenient to decompose the Hamiltonian into a so-called cubic and quadratic component, given respectively by
\begin{align}
\Hthree = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
\Htwo = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}
Then the Hamiltonian can be expressed as
\begin{equation}
\label{Hamiltonian}
\Hupdn = C_2 \dratupdn^{1/2} \, \Htwo - C_3 \dratupdn^{-3/2} \, \Hthree
\end{equation}
where $\drat = \dratupdn$ changes value across the ADC. More explicitly, substituting \eqref{dratpw}, the separate upstream and downstream Hamiltonians are
\begin{align}
&\Hup = C_2 \, \Htwo - C_3 \, \Hthree 						&& \text{for } t<0 \\
&\Hdn = C_2 \dratdn^{1/2} \, \Htwo - C_3 \dratdn^{-3/2} \, \Hthree	&& \text{for } t>0
\end{align}
The above formulation differs cosmetically from some previous ones \cite{abramov2003hamiltonian, bajars2013weakly, majda2019statistical}, which used the opposite sign in both $\sympJ$ and $\Ham$. With that sign convention, Majda et al. 2019 found that a {\em negative} inverse temperature is required to accurately describe the experimental observations \cite{majda2019statistical}. We have chosen the sign convention above so that a {\em positive} inverse temperature could be used and thus fit the theory into the most conventional form of statistical mechanics.

@@

We introduce two important invariants of KdV, the momentum and the energy
\begin{align}
\label{MomEn}
\Mo[u] \equiv \int_{-\pi}^{\pi} u \dx \, = 0 , \qquad
\En[u] \equiv \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = 1
\end{align}
As indicated above, the momentum of $u$ vanishes since it is measured as displacement from equilibrium. Second, by the scale chosen for $\ampscale$ in \eqref{scales}, the energy has been normalized to unity.

@@

Consider the Fourier series of the state variable
\begin{align}
u = \sum_{k=-\Lambda}^{\Lambda} \uhat_k e^{i k x} \, , \qquad
\uhat_k = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x) e^{-i k x} \dx
\end{align}
The above sum can be infinite $\Lambda = \infty$ or truncated at a finite wavenumber $\Lambda < \infty$. Further, since $u$ is real valued, we have $\uhat_{-k} = \uhat_{k}^*$. Since momentum vanishes, we have $\uhat_0 = 0$. Meanwhile, Parseval's identity gives
\begin{equation}
\En[u] = 2 \pi \sum_{k=1}^{\Lambda} \abs{\uhat_k}^2 = 1
\end{equation}
% The above agrees with the first line on top of p. 4 in Majda and Qi J Stat Phys 2019.

The KdV system (i.e. $\Lambda = \infty$) possesses the Hamiltonian \eqref{Hamiltonian} and momentum and energy \eqref{MomEn} and an infinite number of additional invariants \cite{lax1975periodic, whitham2011linear}. For the truncated system however, $\Lambda < \infty$, the only known invariants are the momentum, energy, and Hamiltonian. The infinite sequence of invariants apart from these first three may no longer be invariants of TKdV.
